### 2022-10-18 chapters/introduction.tex
proof read and fix some typos

          In this section, we provide the literature approaches used to measure energy consumption and classify them, compare them and discuss their pros and cons when used for our research. After that, we tackle the challenge of accuracy within the field of energy consumption, and discuss how the literature tried to overcome these challenges. Finally, we  present some of the most recent works in the field of software energy consumption optimization;

>>>>

          In this section, we provide the literature approaches used to measure energy consumption and classify them, compare them and discuss their pros and cons when used for our research. After that, we tackle the challenge of accuracy within the field of energy consumption and discuss how the literature tried to overcome these challenges. Finally, we  present some of the most recent works in the field of software energy consumption optimization;

### 2022-10-18 chapters/optimization/java-dir/java.tex
proof read and fix some typos

% \section*{JAVA}
\section{Introduction}
As reported in the state of the art, Java is one of the most popular programming languages adopted by practitioners.In fact, in 2022, Java will be second only to Python, according to PYPY~\footnote{\url{https://pypl.github.io/PYPL.html}}.
Furthermore, if we take into consideration legacy applications, Java becomes the most used programming language.According to the TIOBE index, Java was the most frequently used language from 2002 until 2017, and it remained in the top 5 after that~\footnote{\url{https://www.tiobe.com/tiobe-index/}}.
In addition to its popularity, Java exhibits an interesting behavior when it comes to energy consumption and performance, Java applications can be at the same time one of the most energy-efficient or hungry solutions.
As we have seen in the previous chapter, an inappropriate combination of parameters can drive Java applications from the top language to the bottom just by setting the wrong parameters.
in this chapter, we want to dig deeper into this aspect of Java and study its runtime.
This chapter thus focuses on the impact of the runtime of Java applications on energy consumption.

@@

Java's portability is a core design goal, which means that Java applications will work exactly the same on any operating system and on any hardware.As we see in figure~\ref{fig:JVM_architecture}, Instead of machine code, this is accomplished by compiling Java language code to an intermediate representation known as Java bytecode. Java bytecode instructions are similar to machine code, but they are executed by a JVM rather than directly on the machine. As a result, Java programs will run slower and consume more memory than C++ programs. However, Just-in-Time (JIT) compilation, which is built into the JVM, improves performance by converting bytecode to machine code at runtime. Furthermore, an automatic garbage collector (GC) is used by Java to manage memory throughout the object lifecycle and to recover memory when objects are no longer in use.

@@

\subsection{Metrics and Measurement}
Since the goal of this study is the green aspect of JVM, our key metric will be the energy consumed by job completed for each JVM configuration.
In addition to the energy consumption, we collected additional metrics to explain the reasons behind the behavior of each experiment.
Those additional metrics are:
\begin{itemize}
    \item execution time,
    \item number of threads.
\end{itemize}

@@

\subsection{Extension}
We also added an extension to the protocol to allow the user to run the same experiment with different configurations.
The package is available in the GitHub repository.\footnote{\url{https://github.com/chakib-belgaid/jvm-comparaison}}
To add extra \textbf{jvm candidates} for the benchmark applications, we added a new configuration file \texttt{jvm.sh} in the root directory of the repository, where we put the name and the version of the jvm to be used.
For the \textbf{input workload},the benchmarks should be provided in the \texttt{benchmarks} directory.
As for extra \textbf{metrics}, one can create a new script file that monitors the experiment and record the metrics.
We provide some examples, such as \texttt{recordpower.sh} to measure the instantaneous power and \texttt{recordthreads.sh} to measure the number of active threads during the experiment.

For faster experiments, we propose \textsc{Jreferral}\footnote{\url{https://github.com/chakib-belgaid/jreferral}}, an open source tool that automatically compares the JVM energy consumption and recommends the most energy-efficient configuration for a given Java application.
We further discuss this tool in Section~\ref{section:SummaryofContributions}.

@@

The second interesting situation was observed on the \textsf{Dotty} benchmark.
More specifically, during the first $100$ seconds of the execution of the \textsf{Dotty} benchmark on all evaluated JVMs.
At the beginning of the execution, \textsc{GraalVM} has a slightly lower power consumption, is faster, and consumes 10\% less energy.
After about $150$ seconds, the power differences between the 3 JVMs is barely noticeable.
One can, however, notice the effect of the JIT, as \textsc{HotSpot} takes the advantage over \textsc{GraalVM} and becomes more energy efficient.
In total, \textsc{HotSpot} completes $597$ requests against $510$ for \textsc{GraalVM} and $381$ for \textsc{J9}, as shown in \Cref{table:power/request}.
\textsc{HotSpot} was thus the best choice in the long term, which explains why it is always necessary to consider a warm-up phase and wait for the JIT to be triggered before evaluating the effect of the JVM or the performance of an application.
This is precisely what we did in our experiments, and it is why \textsc{HotSpo}t was more energy efficient than \textsc{GraalVM} in \Cref{fig:all_benchs}; therefore, ignoring the warm-up phase would have been misleading.

@@

\subsubsection{Multithreading}
The purpose of this phase is to investigate the impact JVM thread management strategies on energy consumption.
This encompasses exploring if the management strategies of application-level parallelism (so-called \emph{threads}) result in different energy efficiencies, depending on JVM distributions.

Investigating such a hypothesis requires a selection of highly parallel and CPU-intensive benchmarks, which is one of the main criteria for our benchmark selection.
As no tool can accurately monitor the energy consumption at a thread level, we monitor the global power consumption and CPU utilization during the execution using RAPL for the energy, and several Linux tools for the CPU-utilization (\texttt{htop}, \texttt{cpufreq}).
%The experiments showed differences in energy consumption between the 3 JVMs and even between 2 version on the HotSpot VM.
Knowing that most of the benchmarks are multi-threaded jobs that use multiple cores, further analysis of thread management is required to understand the results of our previous experiments.
We thus selected the benchmarks that highlighted the highest differences along JVM distributions from \Cref{fig:all_benchs}, namely \textsf{Avrora} and \textsf{Reactors}.
We studied their multi-threaded behavior to optimize their energy efficiency.

@@

\Cref{fig:threads} delivers a closer look to the thread allocation strategies adopted by JVM.
First, \Cref{fig:avrora_threads} illustrates the active threads count evolution over time (excluding the JVM-related threads, usually 1 or 2 extra threads depending on the execution phase) for \textsf{Avrora}.
One can notice through the figure that \textsc{J9} exploits the CPU more intensively by running much more parallel threads compared to other JVMs (an average of 5.1 threads per second for \textsc{J9} while the other JVMs do not exceed 1.5 thread per second).
Furthermore, the number of context switches is twice bigger for \textsc{J9}, while the number of soft page faults is twice smaller.
The efficient \textsc{J9} thread management explains why running the \textsf{Avrora} benchmark took much less time and consumed less energy, given that no other difference for the JIT or GC configuration was spotted between the JVMs.
Another key reason for the \textsc{J9}'s efficiency for the \textsf{Avrora} benchmark is memory allocation, as \textsf{OpenJ9} adopts a different policy for the heap allocation.
It creates a non-collectible \emph{thread local heap} (TLH) within the main heap for each active thread.
The benefit of cloning a dedicated TLH is the fast memory access for independent threads: each thread has its heap and no deadlock can occur.

@@

In conclusion, JVMs thread management can sometimes constitute a key factor that impacts software energy consumption.
However, we suggest checking and comparing JVMs before deploying a software, especially if the target application is parallel and multi-threaded.

@@

For \textsc{GraalVM}, we conducted experiments with and without the JVMCI (a Java-based JVM compiler interface enabling a compiler written in Java to be used by the JVM as a dynamic compiler).
We also considered both the community and economy configurations (no enterprise).
A JIT+AOT (\emph{Ahead Of Time}) disabling experiment has also been considered for all of the 3 JVM platforms.
\Cref{table:JIT} reports on the energy consumption of the experiments we conducted for most of the benchmarks and JIT configurations under study.

@@

For \textsc{J9}, we noticed that adopting the default JIT configuration is always better than specifying a custom JIT intensity.
The \textsf{warm} configuration delivers the closest results to the best results observed with the default configuration.
Moreover, choosing a low minimum count of method calls seems to have a negative effect on the execution time and energy consumption.
The only parameter that can give better performance than the default configuration in some cases is the number of parallel JIT threads---using 3 and 7 parallel threads---but is not statistically significant.

@@

Unfortunately, we cannot convey predictive patterns on how to configure the GC to optimize energy efficiency.
However, some considerations should be taken into account when choosing the GC, such as the garbage collection time, the throughput, etc.
Other settings are less trivial to determine, such as tenure age, memory size, and GC threads count.
Experiments should thus be conducted on the software to tune the most convenient GC configuration to achieve better energy efficiency in production.

@@

\section{Tools and contributions}
\textsf{J-Referral} is an open-source tool designed to assist developers and practitioners in selecting the most energy-efficient JVM configuration for their software.
\Cref{table:j-referral} illustrates an example of the final report returned by \textsf{J-Referral}.
The tool was tested for 2 Java projects: \textsf{Zip4J}\footnote{\url{https://github.com/srikanth-lingala/zip4j}} and \textsf{K-nucleotide}.\footnote{\url{https://benchmarksgame-team.pages.debian.net/benchmarksgame/performance/knucleotide.html}}
\textsf{Zip4J} runs a large file compression, while \textsf{K-nucleotide} extracts a DNA sequence, and updates a hashtable of k-nucleotide keys to count specific values.
The short report presented in \Cref{table:j-referral} shows the ratio of potential energy saving between the most and least energy consuming tested JVM (40\% and 70\% energy savings for \textsf{Zip4J} and \textsf{K-nucleotide}, respectively).
Options are available for \textsf{J-Referral} to obtain much more detailed reports including execution time, DRAM usage, split DRAM vs. CPU consumption, etc.
The tool is available as \emph{open-source software} (OSS) from our GitHub repository.\footnote{\url{https://github.com/chakib-belgaid/jreferral}}

>>>>

% \section*{JAVA}
\section{Introduction}
As reported in the state of the art, Java is one of the most popular programming languages adopted by practitioners. In fact, in 2022, Java will be second only to Python, according to PYPY~\footnote{\url{https://pypl.github.io/PYPL.html}}.
Furthermore, if we take into consideration legacy applications, Java becomes the most used programming language. According to the TIOBE index, Java was the most frequently used language from 2002 until 2017, and it remained in the top 5 after that~\footnote{\url{https://www.tiobe.com/tiobe-index/}}.
In addition to its popularity, Java exhibits an interesting behavior when it comes to energy consumption and performance, Java applications can be at the same time one of the most energy-efficient or hungry solutions.
As we have seen in the previous chapter, an inappropriate combination of parameters can drive Java applications from the top language to the bottom just by setting the wrong parameters.
in this chapter, we want to dig deeper into this aspect of Java and study its runtime.
This chapter thus focuses on the impact of the runtime of Java applications on energy consumption.

@@

Java's portability is a core design goal, which means that Java applications will work the same on any operating system and any hardware. As we see in figure~\ref{fig:JVM_architecture}, Instead of machine code, this is accomplished by compiling Java language code to an intermediate representation known as Java bytecode. Java bytecode instructions are similar to machine code, but they are executed by a JVM rather than directly on the machine. As a result, Java programs will run slower and consume more memory than C++ programs. However, the Just-in-Time (JIT) compilation, which is built into the JVM, improves performance by converting bytecode to machine code at runtime. Furthermore, an automatic garbage collector (GC) is used by Java to manage memory throughout the object lifecycle and to recover memory when objects are no longer in use.

@@

\subsection{Metrics and Measurement}
Since the goal of this study is the green aspect of JVM, our key metric will be the energy consumed by the job completed for each JVM configuration.
In addition to the energy consumption, we collected additional metrics to explain the reasons behind the behavior of each experiment.
Those additional metrics are:
\begin{itemize}
    \item execution time,
    \item number of threads.
\end{itemize}

@@

\subsection{Extension}
We also added an extension to the protocol to allow the user to run the same experiment with different configurations.
The package is available in the GitHub repository.\footnote{\url{https://github.com/chakib-belgaid/jvm-comparaison}}
To add extra \textbf{jvm candidates} for the benchmark applications, we added a new configuration file \texttt{jvm.sh} in the root directory of the repository, where we put the name and the version of the JVM to be used.
For the \textbf{input workload},the benchmarks should be provided in the \texttt{benchmarks} directory.
As for extra \textbf{metrics}, one can create a new script file that monitors the experiment and record the metrics.
We provide some examples, such as \texttt{recordpower.sh} to measure the instantaneous power and \texttt{recordthreads.sh} to measure the number of active threads during the experiment.

For faster experiments, we propose \textsc{Jreferral}\footnote{\url{https://github.com/chakib-belgaid/jreferral}}, an open-source tool that automatically compares the JVM energy consumption and recommends the most energy-efficient configuration for a given Java application.
We further discuss this tool in Section~\ref{section:SummaryofContributions}.

@@

The second interesting situation was observed on the \textsf{Dotty} benchmark.
More specifically, during the first $100$ seconds of the execution of the \textsf{Dotty} benchmark on all evaluated JVMs.
At the beginning of the execution, \textsc{GraalVM} has a slightly lower power consumption, is faster, and consumes 10\% less energy.
After about $150$ seconds, the power differences between the 3 JVMs is barely noticeable.
One can, however, notice the effect of the JIT, as \textsc{HotSpot} takes advantage over \textsc{GraalVM} and becomes more energy efficient.
In total, \textsc{HotSpot} completes $597$ requests against $510$ for \textsc{GraalVM} and $381$ for \textsc{J9}, as shown in \Cref{table:power/request}.
\textsc{HotSpot} was thus the best choice in the long term, which explains why it is always necessary to consider a warm-up phase and wait for the JIT to be triggered before evaluating the effect of the JVM or the performance of an application.
This is precisely what we did in our experiments, and it is why \textsc{HotSpo}t was more energy efficient than \textsc{GraalVM} in \Cref{fig:all_benchs}; therefore, ignoring the warm-up phase would have been misleading.

@@

\subsubsection{Multithreading}
The purpose of this phase is to investigate the impact of JVM thread management strategies on energy consumption.
This encompasses exploring if the management strategies of application-level parallelism (so-called \emph{threads}) result in different energy efficiencies, depending on JVM distributions.

Investigating such a hypothesis requires a selection of highly parallel and CPU-intensive benchmarks, which is one of the main criteria for our benchmark selection.
As no tool can accurately monitor the energy consumption at a thread level, we monitor the global power consumption and CPU utilization during the execution using RAPL for the energy, and several Linux tools for the CPU usage (\texttt{htop}, \texttt{cpufreq}).
%The experiments showed differences in energy consumption between the 3 JVMs and even between 2 version on the HotSpot VM.
Knowing that most of the benchmarks are multi-threaded jobs that use multiple cores, further analysis of thread management is required to understand the results of our previous experiments.
We thus selected the benchmarks that highlighted the highest differences along JVM distributions from \Cref{fig:all_benchs}, namely \textsf{Avrora} and \textsf{Reactors}.
We studied their multi-threaded behavior to optimize their energy efficiency.

@@

\Cref{fig:threads} delivers a closer look at the thread allocation strategies adopted by JVM.
First, \Cref{fig:avrora_threads} illustrates the active threads count evolution over time (excluding the JVM-related threads, usually 1 or 2 extra threads depending on the execution phase) for \textsf{Avrora}.
One can notice through the figure that \textsc{J9} exploits the CPU more intensively by running much more parallel threads compared to other JVMs (an average of 5.1 threads per second for \textsc{J9} while the other JVMs do not exceed 1.5 thread per second).
Furthermore, the number of context switches is twice bigger for \textsc{J9}, while the number of soft page faults is twice smaller.
The efficient \textsc{J9} thread management explains why running the \textsf{Avrora} benchmark took much less time and consumed less energy, given that no other difference for the JIT or GC configuration was spotted between the JVMs.
Another key reason for the \textsc{J9}'s efficiency for the \textsf{Avrora} benchmark is memory allocation, as \textsf{OpenJ9} adopts a different policy for the heap allocation.
It creates a non-collectible \emph{thread local heap} (TLH) within the main heap for each active thread.
The benefit of cloning a dedicated TLH is the fast memory access for independent threads: each thread has its heap and no deadlock can occur.

@@

In conclusion, JVM thread management can sometimes constitute a key factor that impacts software energy consumption.
However, we suggest checking and comparing JVMs before deploying software, especially if the target application is parallel and multi-threaded.

@@

For \textsc{GraalVM}, we conducted experiments with and without the JVMCI (a Java-based JVM compiler interface enabling a compiler written in Java to be used by the JVM as a dynamic compiler).
We also considered both the community and economy configurations (no enterprise).
A JIT+AOT (\emph{Ahead Of Time}) disabling experiment has also been considered for all the 3 JVM platforms.
\Cref{table:JIT} reports on the energy consumption of the experiments we conducted for most of the benchmarks and JIT configurations under study.

@@

For \textsc{J9}, we noticed that adopting the default JIT configuration is always better than specifying a custom JIT intensity.
The \textsf{warm} configuration delivers the closest results to the best results observed with the default configuration.
Moreover, choosing a low minimum count of method calls seems to have a negative effect on execution time and energy consumption.
The only parameter that can give better performance than the default configuration in some cases is the number of parallel JIT threads---using 3 and 7 parallel threads---but is not statistically significant.

@@

Unfortunately, we cannot convey predictive patterns on how to configure the GC to optimize energy efficiency.
However, some considerations should be taken into account when choosing the GC, such as the garbage collection time, the throughput, etc.
Other settings are less trivial to determine, such as tenure age, memory size, and GC thread count.
Experiments should thus be conducted on the software to tune the most convenient GC configuration to achieve better energy efficiency in production.

@@

\section{Tools and contributions}
\textsf{J-Referral} is an open-source tool designed to assist developers and practitioners in selecting the most energy-efficient JVM configuration for their software.
\Cref{table:j-referral} illustrates an example of the final report returned by \textsf{J-Referral}.
The tool was tested for 2 Java projects: \textsf{Zip4J}\footnote{\url{https://github.com/srikanth-lingala/zip4j}} and \textsf{K-nucleotide}.\footnote{\url{https://benchmarksgame-team.pages.debian.net/benchmarksgame/performance/knucleotide.html}}
\textsf{Zip4J} runs a large file compression, while \textsf{K-nucleotide} extracts a DNA sequence, and updates a hashtable of k-nucleotide keys to count specific values.
The short report presented in \Cref{table:j-referral} shows the ratio of potential energy saving between the most and least energy-consuming tested JVM (40\% and 70\% energy savings for \textsf{Zip4J} and \textsf{K-nucleotide}, respectively).
Options are available for \textsf{J-Referral} to obtain much more detailed reports including execution time, DRAM usage, split DRAM vs. CPU consumption, etc.
The tool is available as \emph{open-source software} (OSS) from our GitHub repository.\footnote{\url{https://github.com/chakib-belgaid/jreferral}}

### 2022-10-18 chapters/optimization/python-dir/python.tex
proof read and fix some typos

\subsubsection{Python \& Multiprocessing}
The purpose of this part is to study the impact of concurrency on energy consumption for Python applications.
To do so, we run a simple code snippet that computes the sum of the first $N$ numbers using the standard concurrency libraries \texttt{multiprocessing~\footnote{\url{https://docs.python.org/3/library/multiprocessing.html}}} and \texttt{multihtreading~\footnote{\url{https://docs.python.org/3/library/threading.html}}}.
We run the following strategies using a Desktop machine with Intel(R) Core(TM) i7-4800MQ CPU @ 2.70GHz with four cores and four hyper-threads.
%  N ==5e5
\begin{itemize}
    \item \textsf{Sequential}: we compute the sum of the first $N$ numbers with no concurency;
    \item \textsf{Multithreading}: we use the \texttt{ThreadPoolExecutor} library to compute the sum of the first $N$ numbers. We divide $N$ by the number of threads, and each thread will compute the sum of the numbers in its range. For our case, we used four threads;
    \item \textsf{Multiprocessing}: we use the \texttt{ProcessPoolExecutor} library to compute the sum of the first $N$ numbers. We divide $N$ by the number of processes, and each process sums the numbers in its range. In this situation, we considered 2, 4, 8, and 16 processes iteratively.
\end{itemize}

>>>>

\subsubsection{Python \& Multiprocessing}
The purpose of this part is to study the impact of concurrency on energy consumption for Python applications.
To do so, we run a simple code snippet that computes the sum of the first $N$ numbers using the standard concurrency libraries \texttt{multiprocessing~\footnote{\url{https://docs.python.org/3/library/multiprocessing.html}}} and \texttt{multihtreading~\footnote{\url{https://docs.python.org/3/library/threading.html}}}.
We run the following strategies using a Desktop machine with Intel(R) Core(TM) i7-4800MQ CPU @ 2.70GHz with four cores and four hyper-threads.
%  N ==5e5
\begin{itemize}
    \item \textsf{Sequential}: we compute the sum of the first $N$ numbers with no concurrency;
    \item \textsf{Multithreading}: we use the \texttt{ThreadPoolExecutor} library to compute the sum of the first $N$ numbers. We divide $N$ by the number of threads, and each thread will compute the sum of the numbers in its range. For our case, we used four threads;
    \item \textsf{Multiprocessing}: we use the \texttt{ProcessPoolExecutor} library to compute the sum of the first $N$ numbers. We divide $N$ by the number of processes, and each process sums the numbers in its range. In this situation, we considered the 2, 4, 8, and 16 processes iteratively.
\end{itemize}

### 2022-10-18 chapters/programming_languages/rpc/rpc.tex
proof read and fix some typos

To answer these questions, we will define an interface using the Proto2\footnote{\url{https://developers.google.com/protocol-buffers}} language, and then we will implement it in multiple servers, each one using a different programming language.
We will then compare the energy consumption of each implementation using a custom version of \texttt{GHZ}~\footnote{\url{https://github.com/chakib-belgaid/energy_ghz}}, a tool that allows us to stress test our servers. This fork of \texttt{GHZ} allows us to measure the server's energy consumption during the test's execution and synchronize it with other performance statistics such as Tail latency and the number of requests per second the server can reach.
\begin{figure}[!hbt]
    \begin{center}
        \includegraphics[width=.8\linewidth]{imgs/rpcprotocol}
    \end{center}
    \caption{Experimental software architecture.}\label{fig:rpcprotocol}
\end{figure}
Figure~\ref{fig:rpcprotocol} show the experimental protocol. We will use different messages inserted in the file \texttt{helloworld.proto}.
The rest of this part will be dedicated to explaining each aspect of the experiment.
\subsubsection{Measurement Context}
\paragraph{Hardware settings.}
All the experiments are run on the cluster \textsf{paravance} of the G5K platform.
This cluster comprises 72 identical machines equipped with 2 Intel Xeon E5-2630 V3, with 128\,GB of RAM.
For more accuracy, our SUT (\emph{System Under Test}) runs a minimal version of Debian\,9 (4.9.0 kernel version), which enforces the core processes required for our experiment.
Furthermore, we used Docker containers technology for reproducibility of the experiments and the isolation of the servers.
\paragraph{Client \& server environments.}
To limit networks' impact on the experiments, we run both the client and the server on the same machine.
However, we isolate each part on a separate CPU socket to reduce the noise that the client might have on the server and \emph{vice-versa}.
To do so, for each iteration, we always run the same client on \textsf{socket\,0} and the server that we want to test on \textsf{socket\,1}.
Both the server and the client use the whole socket for their experiment.
In addition, all the additional services, such as the kernel and HwPC sensor, are run on \textsf{socket\,0}.
Therefore, the only process being executed in \textsf{socket\,1} is the server that we benchmark and monitor.

@@

\paragraph{Performances.}
For better accuracy and more details, We updated the open source RPC benchmarking tool GHZ (\url{https://ghz.sh/}).
The modified version allows us to monitor the average power for each request from both the server and the client sides.
The new version is available in the repository.\footnote{\url{https://github.com/chakib-belgaid/energy_ghz}}

@@

\paragraph{Performance Behaviour}
this paragraph will cover the number of requests per second processed per server without looking at its energy.
Here, we consider three observable variables:
\begin{itemize}
    \item \textsf{Satisfaction ratio}: how many requests have been satisfied among the total requests,
    \item \textsf{Request Per Seconds}: The number of the requests that have been answered from the server,
    \item \textsf{Tail Latency at 99\%}: one of the best metrics to evaluate the performances of a server.
\end{itemize}

@@

Figure~\ref{fig:rps_clients} presents the number of requests per second for each implementation. The heatmap is colored in a logarithmic scale to visualize the results better.
As one can observe, most of the servers hit their RPS limit after five clients and 100 clients for VM-based servers, and after this, the number keeps constant, decreasing the average RPS per client.
.Net server is the most performant, followed by Java and Go, while Python and Ruby are the least performant.

@@

For each framework, we can distinguish three modes, and they all depend on the payload size:
\begin{enumerate}
    \item \textsf{Stress-free} mode when the server has enough resources to satisfy the requests, as they require a memory less than a certain threshold (depends on the language and the platform),
    \item \textsf{Escalation} mode when the requests tend to be bigger, but the server can still manage to handle them, at the price of a change in the energy and performance behaviors,
    \item \textsf{Broken state} mode when the size of requests exceeds 1MB, most of the servers cannot handle them, and they tend to crash. After 10MB, except Elixir, no server could handle those requests.
\end{enumerate}

@@

The results show that the number of clients had a binary effect on the servers.
With a small number of clients, there was no typical behavior between different categories.
On the other hand, when dealing with a more significant number of clients, interpreted languages and compiled ones dropped in terms of performance while keeping the same average power, unlike the VM-based ones that increased their average power while keeping the same performances.
Overall, the second strategy had a better impact on the single request cost than the first one.

>>>>

To answer these questions, we will define an interface using the Proto2\footnote{\url{https://developers.google.com/protocol-buffers}} language, and then we will implement it in multiple servers, each one using a different programming language.
We will then compare the energy consumption of each implementation using a custom version of \texttt{GHZ}~\footnote{\url{https://github.com/chakib-belgaid/energy_ghz}}, a tool that allows us to stress test our servers. This fork of \texttt{GHZ} allows us to measure the server's energy consumption during the test's execution and synchronize it with other performance statistics such as Tail latency and the number of requests per second the server can reach.
\begin{figure}[!hbt]
    \begin{center}
        \includegraphics[width=.8\linewidth]{imgs/rpcprotocol}
    \end{center}
    \caption{Experimental software architecture.}\label{fig:rpcprotocol}
\end{figure}
Figure~\ref{fig:rpcprotocol} show the experimental protocol. We will use different messages inserted in the file \texttt{helloworld.proto}.
The rest of this part will be dedicated to explaining each aspect of the experiment.
\subsubsection{Measurement Context}
\paragraph{Hardware settings.}
All the experiments are run on the cluster \textsf{paravance} of the G5K platform.
This cluster comprises 72 identical machines equipped with 2 Intel Xeon E5-2630 V3, with 128\,GB of RAM.
For more accuracy, our SUT (\emph{System Under Test}) runs a minimal version of Debian\,9 (4.9.0 kernel version), which enforces the core processes required for our experiment.
Furthermore, we used Docker containers technology for the reproducibility of the experiments and the isolation of the servers.
\paragraph{Client \& server environments.}
To limit networks' impact on the experiments, we run both the client and the server on the same machine.
However, we isolate each part on a separate CPU socket to reduce the noise that the client might have on the server and \emph{vice-versa}.
To do so, for each iteration, we always run the same client on \textsf{socket\,0} and the server that we want to test on \textsf{socket\,1}.
Both the server and the client use the whole socket for their experiment.
In addition, all the additional services, such as the kernel and HwPC sensor, are run on \textsf{socket\,0}.
Therefore, the only process being executed in \textsf{socket\,1} is the server that we benchmark and monitor.

@@

\paragraph{Performances.}
For better accuracy and more details, We updated the open-source RPC benchmarking tool GHZ (\url{https://ghz.sh/}).
The modified version allows us to monitor the average power for each request from both the server and the client sides.
The new version is available in the repository.\footnote{\url{https://github.com/chakib-belgaid/energy_ghz}}

@@

\paragraph{Performance Behaviour}
this paragraph will cover the number of requests per second processed per server without looking at its energy.
Here, we consider three observable variables:
\begin{itemize}
    \item \textsf{Satisfaction ratio}: how many requests have been satisfied among the total requests,
    \item \textsf{Request Per Seconds}: The number of requests that have been answered from the server,
    \item \textsf{Tail Latency at 99\%}: one of the best metrics to evaluate the performances of a server.
\end{itemize}

@@

Figure~\ref{fig:rps_clients} presents the number of requests per second for each implementation. The heatmap is colored with a logarithmic scale to visualize the results better.
As one can observe, most of the servers hit their RPS limit after five clients and 100 clients for VM-based servers, and after this, the number keeps constant, decreasing the average RPS per client.
.Net server is the most performant, followed by Java and Go, while Python and Ruby are the least performant.

@@

For each framework, we can distinguish three modes, and they all depend on the payload size:
\begin{enumerate}
    \item \textsf{Stress-free} mode when the server has enough resources to satisfy the requests, as they require a memory less than a certain threshold (depends on the language and the platform),
    \item \textsf{Escalation} mode when the requests tend to be bigger, but the server can still manage to handle them, at the price of a change in the energy and performance behaviors,
    \item \textsf{Broken state} mode when the size of requests exceeds 1MB, most of the servers cannot handle them, and they tend to crash. After 10MB, except for Elixir, no server could handle those requests.
\end{enumerate}

@@

The results show that the number of clients had a binary effect on the servers.
With a small number of clients, there was no typical behavior between different categories.
On the other hand, when dealing with a more significant number of clients, interpreted languages and compiled ones dropped in terms of performance while keeping the same average power, unlike the VM-based ones that increased their average power while keeping the same performance.
Overall, the second strategy had a better impact on the single request cost than the first one.
